Formalization of Typed Lua

### 2015-06-12 doc/papers/paper/paper.tex
DLS paper: rewrote explanatin of typing rules for table constructors

Our abstract syntax reduces the syntactic forms of the table constructor
into two forms: $\{\;\overline{[e_{1}] = e_{2}}\;\}$ and
$\{\;\overline{[e_{1}] = e_{2}},me\;\}$.
The first uses a list of table fields $([e_{1}] = e_{2})_{1}, ..., ([e_{1}] = e_{2})_{n}$.
The second uses a list of table fields and an expression that can
produce multiple values.

The simplest expression involving tables in the empty table constructor.
Its type checking rule is straightforward:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR-EMPTY}\\
\env_{1}, \penv \vdash \{\}:\{\}_{unique}, \env_{1}
\end{array}
\]

As a more interesting example, let us see how our type system type checks
the table constructor $\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$.

First, our type system uses the auxiliary relation
$\env_{1}, \penv \vdash [e_{1}] = e_{2} : (K,V), \env_{2}$ to type check each
table field.
This auxiliary relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, checking a table field $[e_{1}] = e_{2}$
produces a pair $(K,V)$ and a new type environment $\env_{2}$.
A pair $(K,V)$ means that $e_{1}$ has type $K$ and $e_{2}$ has type $V$,
where $K$ is the type of the key and $V$ is the type of the value.

After type checking each table field, our type system uses each pair $(K,V)$
to build the table type that express the type of a given constructor, and
uses the predicate \emph{wf} to check whether this table type is well-formed.
Formally, a table type is well-formed if it obeys the following rule:
\[
\forall i \not\exists j \; i \not= j \wedge K_{i} \lesssim K_{j}
\]

Well-formed table types avoid ambiguity.
For instance, this rule detects that the table type
$\{1:\Number, \Integer:\String, \Any:\Boolean\}$ is ambiguous,
because the type of the value stored by key $1$ can be
$\Number$, $\String$, or $\Boolean$, as $1 \lesssim 1$,
$1 \lesssim \Integer$, and $1 \lesssim \Any$.
Moreover, the type of the value stored by a key of type $\Integer$,
which is not the literal type $1$, can be $\Number$ or $\Boolean$,
as $\Integer \lesssim \Integer$, and $\Integer \lesssim \Any$.

Well-formed table types also do not allow \emph{unique} and
\emph{open} table types to appear in the type of the values.
We made this restriction because our type system does not keep
track of aliases to table fields.
This means that allowing \emph{unique} and \emph{open} table
types to appear in the type of a value would allow the
creation of unsafe aliases.
The rules that type check table fields use the auxiliary function
\emph{close} to close the type of the values in a table type.

The rule \textsc{T-CONSTRUCTOR-FIELDS} uses these steps to type check a
table constructor with a non-empty list of table fields:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR-FIELDS}\\
\dfrac{\begin{array}{c}
       \overline{\env_1, \penv \vdash e_2: V, \env_v}
        \;\;\;
       \overline{\env_v, \penv \vdash e_1: K, \env_f}
        \\
       T = \{\overline{K{:}V}\}_{unique} \;\;\;
       \env_{2} = merge(\overline{\env_{f}}) \;\;\; wf(T)
       \end{array}}
      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{2}}
\end{array}
%\dfrac{\begin{array}{c}
%	\env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(K_{i},V_{i}), \env_{i+1} \\
%	T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
%	wf(T) \;\;\;
%	n = |\;\overline{[e_{1}] = e_{2}}\;| \\
%	\env_{f} = merge(\env_{1}, ..., \env_{n+1})
%	\end{array}}
%{\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{f}}
%\end{array}
\]

Back to our example, the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ has type
$\{1:``x", 2:``y", 3:``z"\}_{unique}$ through rule \textsc{T-CONSTRUCTOR-FIELDS}.

As another example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ through rule
\textsc{T-CONSTRUCTOR-FIEDLS}.
The inner table is \emph{closed} to prevent the creation of unsafe aliases.

>>>>

The simplest expression involving tables is the empty
table constructor $\{\}$. The empty table constructor always
has type $\{\}_{unique}$. A more interesting example is $\{ [``x"] = 1, [``y"] = 2 \}$, which has type $\{ ``x":1 , ``y": 2 \}_{unique}$. The subtyping rule for {\em unique} tables lets
us assign this table to a variable with a more general
type such as $\{ ``x":\Number , ``y": \Number \}_{unique}$ or
even $\{ \String:\Number \cup \Nil \}_{unique}$. We give
the typing rule for table constructors of this form below: 
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR-FIELDS}\\
\dfrac{\begin{array}{c}
	\overline{\env_1, \penv \vdash e_2: fix(V), \env_v}
	\;\;\;
	\overline{\env_v, \penv \vdash e_1: K, \env_f}
	\\
	T = \{\overline{K{:}V}\}_{unique} \;\;\;
	\env_{2} = merge(\overline{\env_{f}}) \;\;\; wf(T)
	\end{array}}
{\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{2}}
\end{array}
%\dfrac{\begin{array}{c}
%	\env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(K_{i},V_{i}), \env_{i+1} \\
%	T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
%	wf(T) \;\;\;
%	n = |\;\overline{[e_{1}] = e_{2}}\;| \\
%	\env_{f} = merge(\env_{1}, ..., \env_{n+1})
%	\end{array}}
%{\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{f}}
%\end{array}
\]

Due to the system for changing the table type of variables that
we will explain in this section, any expression can potentially
change the type environment. Lua semantics does not dictate
an order of evaluation among the pairs of field initializers,
so all of them are typed in the same input type environment.
The $merge$ auxiliary function then combines their output
environments, diverging in case of conflicting changes.

The rule also checks if the resulting type is {\em well-formed}
with the predicate $wf$. Formally, we define this predicate
for table types with the following formula:
\[
\begin{array}{c}
\forall i \; ((\nexists j \; i \not= j \,\wedge\, K_{i} \lesssim K_{j}) \,\wedge\, wf(V_{i}) \,\wedge\,\\ \lnot isunique(V_{i}) \,\wedge\, \lnot isopen(V_{i}))
\end{array}
\]

Well-formed table types avoid ambiguity.
For instance, this rule detects that the table type
$\{1:\Number, \Integer:\String, \Any:\Boolean\}$ is ambiguous,
because indexing it with a key of type $\Integer$ could
yield either a number, a string or a boolean at runtime.

Well-formed table types also do not allow \emph{unique} and
\emph{open} table types to appear in the type of the values,
and they must be well-formed. Our type system currently cannot
track aliases of table fields, so this restriction is
necessary.

This restriction is the reason rule {\sc T-CONSTRUCTOR-FIELDS}
uses the auxiliary function $fix$ to change any {\em open}
and {\em unique} table types used in the field initializer
to {\em fixed}. As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
has type $\{``x":1, ``y":\{``z":2\}_{fixed}\}_{unique}$.
